---
title: 'Reproducible Research: Peer Assessment 2'
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

## Synopsis
After loading the data, we only subset the data we need like the data related to population health and property damage with crop damage. Then we sum up the result corresponding to the specific event type. In the damage data set, we need to take the exponent part into consideration. In order to better view the result, we sort the data descending and plot the first 10 result. From the data, we can see that across the United States, TORNADO is the most harmful with respect to population health, while FLOOD causes the most property damage and DRAUGHT causes the most crop damage.

```{r setoptions, echo = TRUE, cache = TRUE}

```

## Data Loading

```{r Process-data, cache = TRUE}
# read csv
stormData <- read.csv("StormData/repdata-data-StormData.csv")
# show names to get a basic understanding 
names(stormData)
```

## Population health
```{r populationHealth}
# subset data for population health
populationHealthData <- stormData[, c("EVTYPE", "FATALITIES", "INJURIES")]
# convert factor to integer for later aggregation
populationHealthData$FATALITIES <- as.integer(populationHealthData$FATALITIES)
populationHealthData$INJURIES <- as.integer(populationHealthData$INJURIES)
# sum up the result and show 
populationHealthResult <- aggregate(cbind(INJURIES, FATALITIES) ~ EVTYPE, data = populationHealthData, sum)
head(populationHealthResult)

# sort the data and show
sortedFatalityResult <- populationHealthResult[order(-populationHealthResult$FATALITIES), ]
head(sortedFatalityResult)

sortedInjuryResult <- populationHealthResult[order(-populationHealthResult$INJURIES), ]
head(sortedInjuryResult)

# plot the sorted result
par(mfrow = c(1,2), mar = c(10, 4, 3, 2))
barplot(sortedFatalityResult$FATALITIES[1:10], names.arg = sortedFatalityResult$EVTYPE[1:10], las = 3, ylab = "Fatalities", main = "10 greatest causes of fatalities")
barplot(sortedInjuryResult$INJURIES[1:10], names.arg = sortedInjuryResult$EVTYPE[1:10], las = 3, ylab = "Injuries", main = "10 greatest causes of injuries")
```

From the plot, we can see that across the United States, TORNADO is the most harmful with respect to population health.

## Economic consequences
```{r}
# show names
names(stormData)
# subset data for economic consequences
economicData <- stormData[, c("EVTYPE", "PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP")]
# convert factor to integer for later aggregation
economicData$PROPDMG <- as.numeric(economicData$PROPDMG)
economicData$CROPDMG <- as.numeric(economicData$CROPDMG)

head(economicData)
# deal with exponent
unique(economicData$PROPDMGEXP)

economicData$PROPEXP[economicData$PROPDMGEXP == "K"] <- 1000
economicData$PROPEXP[economicData$PROPDMGEXP == "M"] <- 1e+06
economicData$PROPEXP[economicData$PROPDMGEXP == ""] <- 1
economicData$PROPEXP[economicData$PROPDMGEXP == "B"] <- 1e+09
economicData$PROPEXP[economicData$PROPDMGEXP == "m"] <- 1e+06
economicData$PROPEXP[economicData$PROPDMGEXP == "+"] <- 0
economicData$PROPEXP[economicData$PROPDMGEXP == "0"] <- 1
economicData$PROPEXP[economicData$PROPDMGEXP == "5"] <- 1e+05
economicData$PROPEXP[economicData$PROPDMGEXP == "6"] <- 1e+06
economicData$PROPEXP[economicData$PROPDMGEXP == "?"] <- 0
economicData$PROPEXP[economicData$PROPDMGEXP == "4"] <- 1e+04
economicData$PROPEXP[economicData$PROPDMGEXP == "2"] <- 1e+02
economicData$PROPEXP[economicData$PROPDMGEXP == "3"] <- 1e+03
economicData$PROPEXP[economicData$PROPDMGEXP == "h"] <- 1e+02
economicData$PROPEXP[economicData$PROPDMGEXP == "7"] <- 1e+07
economicData$PROPEXP[economicData$PROPDMGEXP == "H"] <- 1e+02
economicData$PROPEXP[economicData$PROPDMGEXP == "-"] <- 0
economicData$PROPEXP[economicData$PROPDMGEXP == "1"] <- 1e+01
economicData$PROPEXP[economicData$PROPDMGEXP == "8"] <- 1e+08
    
unique(economicData$CROPDMGEXP)

economicData$CROPEXP[economicData$CROPDMGEXP == ""] <- 1
economicData$CROPEXP[economicData$CROPDMGEXP == "M"] <- 1e+06
economicData$CROPEXP[economicData$CROPDMGEXP == "k"] <- 1e+03
economicData$CROPEXP[economicData$CROPDMGEXP == "m"] <- 1e+06
economicData$CROPEXP[economicData$CROPDMGEXP == "B"] <- 1e+09
economicData$CROPEXP[economicData$CROPDMGEXP == "?"] <- 0
economicData$CROPEXP[economicData$CROPDMGEXP == "0"] <- 1
economicData$CROPEXP[economicData$CROPDMGEXP == "k"] <- 1e+03
economicData$CROPEXP[economicData$CROPDMGEXP == "2"] <- 1e+02

# Compute the value
economicData$PROPDMGVALUE <- economicData$PROPDMG * economicData$PROPEXP
economicData$CROPDMGVALUE <- economicData$CROPDMG * economicData$CROPEXP

head(economicData)
# sum up the result and show 
economicResult <- aggregate(cbind(PROPDMGVALUE, CROPDMGVALUE) ~ EVTYPE, data = economicData, sum)
head(economicResult)

# sort the data and show
sortedPropResult <- economicResult[order(-economicResult$PROPDMGVALUE), ]
head(sortedPropResult)

sortedCropResult <- economicResult[order(-economicResult$CROPDMGVALUE), ]
head(sortedCropResult)
# plot the sorted result
par(mfrow = c(1,2), mar = c(10, 4, 3, 2))
barplot(sortedPropResult$PROPDMGVALUE[1:10], names.arg = sortedPropResult$EVTYPE[1:10], las = 3, ylab = "PROPDMG", main = "10 greatest causes of PROPDMG")
barplot(sortedCropResult$CROPDMGVALUE[1:10], names.arg = sortedCropResult$EVTYPE[1:10], las = 3, ylab = "CROPDMG", main = "10 greatest causes of CROPDMG")
```

From the plot, we can see FLOOD causes the most property damage and DRAUGHT causes the most crop damage.
